A Physically Motivated Framework to Compare Merger Timescales\\ of Isolated Low- and High-Mass Galaxy Pairs Across Cosmic Time

### 2024-03-31 paper/main.tex
Updates from Overleaf

% defining affiliations
\newcommand{\affuofa}{University of Arizona, 933 N. Cherry Ave,
    Tucson, AZ 85721, USA}
\newcommand{\affuofu}{Department of Astronomy, University of Utah, Salt Lake City, UT 84112, USA}

@@

\begin{abstract}
    % Statement
    The merger timescales of isolated low-mass pairs ($\rm 10^8<M_*<5\times10^9\,\Msun$) on cosmologically motivated orbits have not been studied in detail, though isolated high-mass pairs ($\rm 5\times10^9<M_*<10^{11}\,\Msun$) have been studied extensively.
    It is common to apply the same separation criteria and expected merger timescales of high-mass pairs to low-mass systems, however it is unclear if their merger timescales are similar, nor if they evolve similarly with redshift. 
    % Our solution
    We use the Illustris TNG100 simulation to quantify the merger timescales of isolated low-mass and high-mass major pairs as a function of cosmic time, and explore how different selection criteria impact the mass and redshift dependence of merger timescales. 
    % What are the timescales? 
    We find that the merger timescales for low- and high-mass pairs are the same from $z=0-6$ in the absence of any separation criteria.
    We then present a physically-motivated framework for selecting pairs via a scaled separation criteria, wherein pair separations are scaled by the virial radius of the primary's FoF group halo ($r_{\mathrm{sep}}< 1 \Rvir$).  
    Applying these scaled separation criteria recovers equivalent merger timescales for both mass scales at all redshifts.
    Alternatively, static physical separation selections applied equivalently to all galaxy pairs at all redshifts leads a difference in merger rates of up to $\sim1\Gyr$ between low- and high-mass pairs, particularly for $\rsep<150\kpc$. 
    % Importance
    As a result, applying the same merger timescales to physical separation-selected pairs will lead to a bias that systematically over-predicts low-mass galaxy merger rates. 
    % could replace the above line with something softer suggesting the application of scaled selection criteria for studies of merger properties of pairs at different mass scales and redshifts, or something like the below:    
    % Our findings suggest that the comparison of low-mass and high-mass galaxy pair fractions, merger timescales, and merger rates is self-consistent using separation criteria that scales as a function of mass and redshift of the system. 
    % if using the same separation criteria to select low-mass galaxy pairs for such studies, or if using the same separation criteria at different redshifts, permits an equitable comparison between galaxies of different mass and redshift.

@@

In \chambe{}, 
abundance matching prescription was sampled 1000 times for each subhalo to account for the spread in the Stellar Mass -- Halo Mass (SMHM) relation. 
For the present study, we only use the stellar masses given by the median of the abundance matching relationship. 
We expect that the spread of merger timescales from different orbital configurations, as well as the redshift spacing of the TNG100 snapshots, will dominate over uncertainties from the number of pairs in the catalog, which varied by $\sim3\%$ (as shown in Fig.~1 in \chambe{}). 
% Since we will be following the orbits of selected pairs both backwards and forwards in time, the

Primary subhalos are defined as the subhalo with the highest assigned stellar mass, $M_{*1}$ in the FoF group, and secondaries are defined as the second most massive subhalo with stellar mass $M_{*2}$. 
Our sample of major pairs then consists of all pairs of primary and secondary subhalos with 
\begin{align*} 
\mbox{\textbf{low mass primaries:}}&\, 10^{8}< M_{*1} < 5\times10^{9} \Msun \\ 
\mbox{\textbf{high mass primaries:}}&\, 5\times 10^{9}< M_{*1} < 10^{11} \Msun\\
\mbox{\textbf{stellar mass ratio:}}&\,      
    M_{*2}/M_{*1} > 1/4.
\end{align*}
A primary or secondary subhalo can only be a member of one single pair at a given snapshot, such that a collection of $N$ pairs consists of $N$ unique primaries and $N$ unique secondaries.
A subhalo can belong to different pairs at different redshifts. 
For example, the primary of a pair that merges at $z=2$ can be selected with a different secondary at $z=1$, constituting a new pair. 
More detail regarding the uniqueness of pairs and orbits is discussed in~\ref{sec:methods-unique}.

@@

\subsection{Mergers} \label{subsec:mergers}
Here we describe how mergers are identified. 
Note that this definition is specific to subhalo mergers and therefore results are not guaranteed to hold for galaxy mergers~\citep[see e.g.,][]{RG2015,Patton2024}, especially in cases where dark matter subhalos and galaxies have different centers of mass, as is the case in our own Galaxy~\citep{Gomez2015,GaravitoCamargo2019,Petersen2021,Chamberlain2023}. % might  note that this is discussed in more deets in discussion - I think Patton 2024 addresses this a bit?
The primary and secondary subhalos of each pair have a `SubfindID' that is used to identify them in the \subfind{} merger tree catalogs, which contain information about every FoF group and subhalo at each redshift. 
The \sublink{} catalogs track subhalos from one redshift to the next, and thus enable us to track subhalos both backwards and forwards in time from any given redshift.

We utilize the \texttt{DescendantID} field of the \sublink{} catalogs to determine which pairs from our base sample merge before the end of the simulation (at $z=0$) and when each merger occurs. 
The \texttt{DescendantID} field provides the `SubhaloID'\footnote{Note that the `SubhaloID' is distinct from the `SubfindID', and is unique for every subhalo in the merger trees. 
The \sublink{} catalogs provide the associated `SubfindID' of each subhalo.} of the subhalo's descendent in the next (or one of the following) snapshots, if it has one.

@@

For each merging pair, we define the `merger redshift' as the redshift that immediately follows the first snapshot where the primary and secondary have the same \texttt{DescendantID}. 
For example, if the primary and secondary have different \texttt{DescendantIDs} from $z=6$ to $z=2$, but have the same \texttt{DescendantID} at $z=2$, then the merger must take place some time between $z=2$ and $z=1.9$, which is the next redshift corresponding to a snapshot in the simulation. 
In this example, we take $z=1.9$ to be the merger redshift.

@@

We extract orbits for all mergers and non-mergers in our base pair sample. 
An orbit for a single pair is defined to be the physical separation between the primary and secondary subhalo as a function of redshift (or lookback time).

@@

In addition to the physical separation, we also calculate the scaled separation of a pair at each redshift. We use the definition of scaled separation $\rsep/\Rvir$ from~\cite{Chamberlain2024}, where $\rsep$ is the physical separation in $\kpc$, and $\Rvir$ is the virial radius of the pair's FoF group (also in $\kpc$).\footnote{Given by \texttt{Group\_R\_TopHat200} in the group catalogs. If the secondary is not in the same FoF group as the primary, the virial radius used to calculate the scaled separation will remain that of the primary's FoF group, which merging secondaries will eventually re-enter prior to merger.} 
The virial radius of the pair's FoF group reasonably approximates the virial radius of a halo with a virial mass equal to the combined subhalo mass of the primary and secondary.
This ``scaled separation" is, by construction, a function of mass and redshift.
%, which will account for the mass difference between low-mass and high-mass pairs and for halo growth over time.

@@

For some pairs, the secondary subhalos can experience first infall, then after one pericentric passage can return to large distances where they are temporarily assigned a different FoF group halo than that of their primary. 
In these cases, we use the full orbit beginning at first infall, including the sections of the orbit where the secondary is assigned to another FoF group temporarily,  through to merger as described in Section \ref{subsec:mergers}.
This definition is more robust for considering the full interaction timescales of merging pairs.
%the post-infall orbit definition is more robust for considering the full interaction timescales of merging pairs than considering the orbit only when the FoF group of the primary and secondary are the same. }
% a variety of different orbits

The middle panel of Fig.~\ref{fig:example-orbits} (``Potential Future Mergers") shows the orbits for three pairs that are classified as non-mergers, since they do not merge before $z=0$, but which appear likely to merge within a few Gyr past the end of the simulation. 
These orbits can have a variety of orbital periods, and the number of pericentric passages can vary significantly. 
The two lighter blue orbits are very long period orbits with 1-2 pericenter passages in the past $10\,\Gyr$, while the darker blue orbit has a much shorter period with many close encounters throughout, and three close passages in just the past $2\,\Gyr$.

The bottom panel shows the orbits of ``Fly-by" interactions (non-mergers) that are unlikely to merge in the near future, if ever.  
Note that we do not split our non-merger category into fly-bys and potential future mergers for any of our following analyses, and such distinctions were made only for the purposes of showing the diversity of selected orbits.

\subsubsection{Uniqueness of orbits}
\label{sec:methods-unique}
Since we collect the orbit (the separation at every redshift where the primary and secondary both exist) for each pair at all redshifts after infall, a singular orbit will be selected as many times as the number of redshifts (or snapshots) where that pair exists. However, we only want to keep a single instance of any given orbit in our catalog, since the orbits with the longest timescales will be selected many times which would skew our data to longer merger timescales.

To distinguish the collection of unique orbits, each pair is assigned a `pairkey' while constructing their orbit. 
The pairkey is created by concatenating the earliest SubhaloID of the primary and secondary subhalo from the \sublink{} catalogs, and is unique for each pair of halos. 
After each pair is assigned a unique pairkey, we only keep one instance of an orbit per pair to avoid double/multi-counting in our orbit catalog.

Note that a single subhalo may be a member of many different pairs, but will only have one unique orbit per unique pair.
For example, the primary of a low-mass pair selected at $z=3$ that merges before $z=2$ may be selected via our selection criteria again at $z=1$ with a new secondary companion. 
In this case, both orbits (of the original low-mass pair and the new pair which includes the primary subhalo from the previous merger) are retained in the orbit catalog, since each orbit is unique.

The total number of pairs (including double counting the same pair at multiple redshifts) is 71,429 for low-mass pairs, and 20,824 high-mass pairs. 
However, after removing all redundant orbits, there remain 22,213 low-mass orbits, and 3,029 high-mass orbits that each correspond to a unique pair of subhalos.

The collection of all unique orbits constitutes our orbit sample and is the dataset that will be used for the remainder of the paper.

@@

\begin{figure}[htb]
    \begin{center}
    \includegraphics[width=\columnwidth]{plots/bet-on-it/6_mergerfraction.png}
    \caption{The fraction of low-mass (green) and high-mass (pink) post-infall orbits that result in a merger before z=0 as a function of redshift. 
    The horizontal line shows a merger fraction of 0.9, or 90\%. 
    High-mass pairs have merger fractions greater than 0.9 for $z>1.1$, and low-mass pairs for $z>1.6$.
    %At $z>1$, the fraction of selected pairs that merge before $z=0$ is upwards of 80\% for both low mass and high mass isolated pairs. 
    The sharp decline of the merger fraction to zero at low redshift is a non-physical feature of the simulation ending at $z=0$.
    Only orbits with sufficiently short merger timescales will be able to merge at these redshifts.}
    \label{fig:fmerge}
    \end{center}
\end{figure}

\subsection{Number of Pairs}\label{sec:pairprops-num}
We calculate the total number of pairs by selecting all pairs at a given redshift, including merger and non-merger pairs.
A single, non-merging pair with first infall at $z=3$ will contribute to the number of pairs at all redshifts from $z=0-3$.

@@

As subhalos experience first infall into a FoF group, the number of pairs increases. 
On the other hand, mergers cause a decrease in the number of pairs.
% The difference between high and low mass
The number of low-mass pairs increases from $\sim600$ pairs at $z=6$ to $\sim6,600$ at $z=2$. 
The decrease in the number of pairs after $z=2$ means that the number of low-mass pairs that are merging at each redshift is larger than the rate at which pairs are being added to the sample.

@@

\subsection{Merger Fraction}\label{sec:pairprops-frac}
    We calculate the merger fraction by dividing the number of pairs that merge (dashed lines in Fig.~\ref{fig:numorbits}) by the total number of merging and non-merging orbits (solid lines) at a given redshift. 
    As before, an orbit with first infall at $z=2$ and merger at $z=1$ will be included in the merger fraction calculation for redshifts $z=1-2$.
    Note, this definition of the merger fraction differs from that typically used in observational studies, where merger fractions are computed from close pair fractions with a correction term.

@@

\begin{figure*}[htb!]
    \begin{center}
    \includegraphics[width=0.7\textwidth]{plots/bet-on-it/8_2Dhist.png}
    \caption{The distribution of merger timescales for low-mass (top) and high-mass (bottom) pairs as a function of physical separation at $z=0.5,1,1.5,2,3,4,5\mbox{and }6$. 
    The colorbars show the percentage of the total number of pairs at each redshift (given in the bottom right corner of each panel) that are in a given bin.
    In each of the first panels, the grey horizontal dashed line shows the time remaining in the simulation, above which there are no mergers. 
    %
    We find:
    1.) a positive correlation between separation and the time until merger at each selected redshift -- the lowest separation pairs tend to have the shortest merger timescales; 
    2.) given a redshift and separation, the low-mass merger timescale is longer than the high-mass merger timescale, i.e. low-mass pairs at $z=1$ with separations $\rsep<100\,\kpc$ merge within $6\, \Gyr$, while high-mass pairs in the same separation range merge within $3\, \Gyr$; 
    3.) merger timescales are longer for lower redshift pairs; 
    4.) the spread of merger timescales and separations increase with decreasing redshift.
    %
    This plot can be used to get estimates for the merger timescale of an isolated pair at a given redshift. 
    For example, a low-mass pair at $z=2$ with $\rsep\sim 75\,\kpc$ will merge in $0-5\,\Gyr$, with a most likely time to merger of around $2\Gyr$.
    On the other hand, a high-mass pair at $z=2$ with $\rsep\sim 75\,\kpc$ will merge in $0-2\,\Gyr$, with a most likely merger timescale of around $0-1\Gyr$.
    }
    \label{fig:2dhist}
    \end{center}
\end{figure*}

\section{Results: The Mass and Redshift Dependence of Merger Timescales}\label{sec:results}
    Using our sample of isolated low-mass and high-mass pairs in the TNG100 simulation, we calculate the merger timescales, or time until merger, for all of the merging pairs in our catalog. 
    The merger timescale of a pair is defined to be the amount of time that elapses between the redshift at which pairs are selected and the merger redshift.
    Only the merging pairs will be considered for the remainder of the analysis. \footnote{Since our orbits are only defined after infall, the merger timescales will likewise only be calculated for pairs that are post-infall at the given separation.}
    Note that we choose not to interpolate the orbits to get a more precise determination of the merger timescale, as the spread on the merger timescales in our study will be dominated by the variation of orbital configurations, rather than the uncertainty due to snapshot spacing. However,~\citet{Patton2024} recently showed that for studies needing more precise determination of, e.g., pericenter passages or merger times, their 6D interpolation scheme can improve timing uncertainties from $\pm80\Myr$ to $\pm3.3\Myr$.

@@

    We find that the time until merger for low-mass and high-mass pairs is positively correlated with the separation of the pair at each redshift. 
    Pairs at $z=4$ with separations between $50-100\,\kpc$ tend to merge in less time than pairs with the same separations at $z=0.5$. 
    For the same physical separation, the merger timescale of high-mass pairs is shorter than for low-mass at each redshift. 
    % However, the slope of this correlation decreases at higher redshifts.

    In addition, the spreads in the merger timescales and pair separations are smaller at higher redshift, for both low- and high-mass pairs. 
    For example, the spread of merger timescales for low-mass pairs at $z=1$ goes from $0-8\,\Gyr$, while pairs at $z=4$ have merger timescales between $0-3\,\Gyr$. 
    Likewise, the spread of separations at $z=1$ is $0-200\kpc$, but at $z=4$ the spread is smaller, between $10-125\kpc$. 
    The high-mass pairs also have a larger spread in the distribution of merger timescales and separations at low redshift than at higher redshift.
    This is due to the growth of halos over time, and thus an increasing virial radius for FoF groups at lower redshifts.

@@

    \subsubsection{Full sample}
        We calculate the merger timescales for pairs at each redshift in the simulation from $z=0-6$, and quantify the median and spread on the merger timescale as a function of redshift. 
        We include the full catalog of merging orbits at each redshift, only excluding separations $\rsep<10\kpc$ to limit the impact of subhalos becoming indistinguishable in the \subfind{} catalogs.% to calculate the merger timescale. 
        % why do we want to consider the entire population of halos? why interesting?

@@

        Figure~\ref{fig:timescales} presents the median merger timescales for merging pairs from $z=0-6$. 
        The 1st and 3rd quartiles are indicated with shaded regions. 
        % low vs high
        Low-mass merger timescales (green) and high-mass merger timescales (pink) are roughly equivalent at all redshifts, which implies that mergers do not proceed in fundamentally different ways at different mass scales. 
        We explore this point further in Sections~\ref{sec:results-phys} and~\ref{sec:results-scal}.

@@

        It is notable that such a simple scaling holds for a complex simulation like TNG100 where subhalos are far from isotropic spherical distributions of mass.
        It is particularly interesting as well that low-mass and high-mass pairs have the same merger timescales \textit{and} follow the $1/H(z)$ redshift scaling. 
        We believe this is due to the dominance of dynamical friction in determining the redshift behavior of merger timescales. 
        % Thus, the primary mechanics that determine merger timescales are likely dynamical friction

@@

    \subsubsection{Physical Separation Selected Pairs }
    \label{sec:results-phys}
        Since pair samples are typically picked via separation criteria in both simulations and observations, we study the impact of different separation-based selection criteria on inferred merger timescales.
        Separation selected samples have a lower separation criteria of $10\,\kpc$.
        This lower separation criteria is also commonly applied to observationally selected pairs in studies of merger fractions and merger rates~\citep[see e.g.][and observational studies cited therein]{Lotz2011,Snyder2017,Besla2018,Chamberlain2024}.

>>>>

% defining affiliations
\newcommand{\affuofa}{University of Arizona, 933 N. Cherry Ave,
    Tucson, AZ 85721, USA}
\newcommand{\affuofu}{Department of Physics and Astronomy, University of Utah, 115 South 1400 East, Salt Lake City, Utah 84112, USA}

@@

\begin{abstract}
    % Statement
    The merger timescales of isolated low-mass pairs ($\rm 10^8<M_*<5\times10^9\,\Msun$) on cosmologically motivated orbits have not yet been studied in detail, though isolated high-mass pairs ($\rm 5\times10^9<M_*<10^{11}\,\Msun$) have been studied extensively.
    % While the merger timescales of isolated high-mass pairs ($\rm 5\times10^9<M_*<10^{11}\,\Msun$) in cosmological environments have been studied extensively, isolated low-mass pairs ($\rm 10^8<M_*<5\times10^9\,\Msun$) have not yet been studied in detail.
    It is common to apply the same separation criteria and expected merger timescales of high-mass pairs to low-mass systems, however it is unclear if their merger timescales are similar, nor if they evolve similarly with redshift. 
    % Our solution
    We use the Illustris TNG100 simulation to quantify the merger timescales of isolated low-mass and high-mass major pairs as a function of cosmic time, and explore how different selection criteria impact the mass and redshift dependence of merger timescales. 
    % What are the timescales? 
    We find that the merger timescales for low- and high-mass pairs are the same from $z=0-6$ in the absence of any separation criteria.
    We then present a physically-motivated framework for selecting pairs via a scaled separation criteria, wherein pair separations are scaled by the virial radius of the primary's FoF group halo ($r_{\mathrm{sep}}< 1 \Rvir$).  
    Applying these scaled separation criteria recovers equivalent merger timescales for both mass scales at all redshifts.
    Alternatively, static physical separation selections applied equivalently to all galaxy pairs at all redshifts leads to a difference in merger rates of up to $\sim1\Gyr$ between low- and high-mass pairs, particularly for $\rsep<150\kpc$. 
    % Importance
    As a result, applying the same merger timescales to physical separation-selected pairs will lead to a bias that systematically over-predicts low-mass galaxy merger rates. 
    % could replace the above line with something softer suggesting the application of scaled selection criteria for studies of merger properties of pairs at different mass scales and redshifts, or something like the below:    
    % Our findings suggest that the comparison of low-mass and high-mass galaxy pair fractions, merger timescales, and merger rates is self-consistent using separation criteria that scales as a function of mass and redshift of the system. 
    % if using the same separation criteria to select low-mass galaxy pairs for such studies, or if using the same separation criteria at different redshifts, permits an equitable comparison between galaxies of different mass and redshift.

@@

In \chambe{}, the
abundance matching prescription was sampled 1000 times for each subhalo to account for the spread in the Stellar Mass -- Halo Mass (SMHM) relation. 
For the present study, we only use the stellar masses given by the median of the abundance matching relationship. 
We expect that the spread of merger timescales from different orbital configurations, as well as the redshift spacing of the TNG100 snapshots, will dominate over uncertainties from the number of pairs in the catalog, which varied by $\sim3\%$ (as shown in Fig.~1 in \chambe{}). 
% Since we will be following the orbits of selected pairs both backwards and forwards in time, the

Primary subhalos are defined as the subhalo with the highest assigned stellar mass, $M_{*1}$ in the FoF group, and secondaries are defined as the second most massive subhalo with stellar mass $M_{*2}$. 
Our sample of major pairs then consists of all pairs of primary and secondary subhalos with 
\begin{align*} 
\mbox{\textbf{low mass primaries:}}&\, 10^{8}< M_{*1} < 5\times10^{9} \Msun \\ 
\mbox{\textbf{high mass primaries:}}&\, 5\times 10^{9}< M_{*1} < 10^{11} \Msun\\
\mbox{\textbf{stellar mass ratio:}}&\,      
    M_{*2}/M_{*1} > 1/4.
\end{align*}
A primary or secondary subhalo can only be a member of one single pair at a given snapshot, such that a collection of $N$ pairs consists of $N$ unique primaries and $N$ unique secondaries.
A subhalo can belong to multiple different pairs at different redshifts. 
For example, the primary of a pair that merges at $z=2$ can be selected with a different secondary at $z=1$, constituting a new pair. 
More detail regarding the uniqueness of pairs and orbits is discussed in Sec.~\ref{sec:methods-unique}.

@@

\subsection{Mergers} \label{subsec:mergers}
Here we describe how mergers are identified. 
Note that this definition is specific to subhalo mergers and therefore results are not guaranteed to hold for galaxy mergers~\citep[see e.g.,][]{RG2015,Patton2024}, especially in cases where dark matter subhalos and galaxies have different centers of mass, as is the case in our own Galaxy~\citep{Gomez2015,GaravitoCamargo2019,Petersen2021,Chamberlain2023}. % might  note that this is discussed in more deets in discussion - I think Patton 2024 addresses this a bit?

The primary and secondary subhalos of each pair have a \texttt{SubfindID} in the \subfind{} merger tree catalogs, which contain information about every FoF group and subhalo at each redshift. 
The \sublink{} catalogs track subhalos from one redshift to the next, and thus enable us to track subhalos both backwards and forwards in time from any given redshift.

We utilize the \texttt{DescendantID} field of the \sublink{} catalogs to determine which pairs from our base sample merge before the end of the simulation (at $z=0$) and when each merger occurs. 
The \texttt{DescendantID} field provides the \texttt{SubhaloID}\footnote{Note that the \texttt{SubhaloID} is distinct from the \texttt{SubfindID}, and is unique for every subhalo in the merger trees. 
The \sublink{} catalogs provide the associated \texttt{SubfindID} of each subhalo.} of the subhalo's descendent in the next (or one of the following) snapshots, if it has one.

@@

For each merging pair, we define the \emph{merger redshift} as the redshift that immediately follows the first snapshot where the primary and secondary have the same \texttt{DescendantID}. 
For example, if the primary and secondary have different \texttt{DescendantIDs} from $z=6$ to $z=2$, but have the same \texttt{DescendantID} at $z=2$, then the merger must take place between $z=2$ and $z=1.9$, which is the next redshift corresponding to a snapshot in the simulation. 
In this example, we take $z=1.9$ to be the merger redshift.

@@

We extract orbits for all mergers and non-mergers in our base pair sample using the merger tree catalogs. 
An orbit for a single pair is defined to be the physical separation between the primary and secondary subhalo as a function of redshift (or lookback time).

@@

In addition to the physical separation, we also calculate the scaled separation of a pair at each redshift. We use the definition of scaled separation $\rsep/\Rvir$ from~\cite{Chamberlain2024}, where $\rsep$ is the physical separation in $\kpc$, and $\Rvir$ is the virial radius of the pair's FoF group.\footnote{Given by \texttt{Group\_R\_TopHat200} in the group catalogs. If the secondary is not in the same FoF group as the primary, the virial radius used to calculate the scaled separation will remain that of the primary's FoF group, which merging secondaries will eventually re-enter prior to merger.} 
The virial radius of the pair's FoF group reasonably approximates the virial radius of a halo with a virial mass equal to the combined subhalo mass of the primary and secondary.
This ``scaled separation" is, by construction, a function of mass and redshift.
%, which will account for the mass difference between low-mass and high-mass pairs and for halo growth over time.

@@

For some pairs, the secondary subhalos can experience first infall, then after one pericentric passage can return to large distances where they are temporarily assigned a different FoF group halo than that of their primary. 
In these cases, we use the full orbit beginning at first infall, including the segments of the orbit where the secondary is assigned to another FoF group temporarily,  through to merger as described in Section \ref{subsec:mergers}.
This definition is most robust for considering the full interaction timescales of merging pairs.
%the post-infall orbit definition is more robust for considering the full interaction timescales of merging pairs than considering the orbit only when the FoF group of the primary and secondary are the same. }
% a variety of different orbits

The middle panel of Fig.~\ref{fig:example-orbits} (``Potential Future Mergers") shows the orbits for three pairs that are classified as non-mergers, since they do not merge before $z=0$, but which appear likely to merge within a few Gyr past the end of the simulation. 
These orbits can have a variety of orbital periods, and the number of pericentric passages can vary significantly. 
The two lighter blue orbits are very long period orbits with 1-2 pericenter passages in the past $10\,\Gyr$, while the darker blue orbit has a much shorter period with three close passages in just the past $2\,\Gyr$.

The bottom panel shows the orbits of ``Fly-by" interactions (non-mergers) that are unlikely to merge in the near future, if ever.  
Note that we do not split our non-merger category into fly-bys and potential future mergers for any of our following analyses, and such distinctions were made only for the purposes of showing the diversity of selected orbits. We further discuss the role of non-mergers and fly-bys on our results in Sec.~\ref{sec:disc-nonmergers}.

\subsubsection{Uniqueness of orbits}
\label{sec:methods-unique}
Since we collect the orbit for each pair at all redshifts after infall, a singular orbit will be selected as many times as the number of redshifts (or snapshots) where that pair exists. However, it is only necessary to keep a single instance of any given orbit in our catalog to avoid skewing our data artificially to longer merger timescales.
%since the orbits with the longest timescales will be selected many times which would skew our data to longer merger timescales.

To distinguish the collection of unique orbits, each pair is assigned a `pairkey' while constructing their orbit. 
The pairkey is created by concatenating the earliest \texttt{SubhaloID} of the primary and secondary subhalo from the \sublink{} catalogs, and is unique for each pair of halos. 
After each pair is assigned a unique pairkey, we only keep one instance of an orbit per pair to avoid double/multi-counting in our orbit catalog.

Note that a single subhalo may be a member of many different pairs, but will only have one unique orbit per unique pair.
For example, the primary of a low-mass pair selected at $z=3$ that merges before $z=2$ may be selected via our selection criteria again at $z=1$ with a new secondary companion. 
In this case, both orbits (of the original low-mass pair and the new pair which includes the primary subhalo from the previous merger) are retained in the orbit catalog.%, since each orbit is unique.

The total number of orbits, before removing the redundant orbits,  %(including double counting the same pair at multiple redshifts) 
is 71,429 for low-mass pairs, and 20,824 high-mass pairs. 
However, after removing all redundant orbits, there remain 22,213 low-mass orbits, and 3,029 high-mass orbits that each correspond to a unique pair of subhalos. The collection of all unique orbits constitutes our orbit sample and is the dataset that will be used for the remainder of this analysis.

Note that we choose not to interpolate the orbits to get a more precise determination of the merger timescale, as the spread on the merger timescales in our study will be dominated by the variation of orbital configurations, rather than the uncertainty due to snapshot spacing. However,~\citet{Patton2024} recently showed that for studies needing more precise determination of, e.g., pericenter passages or merger times, their 6D interpolation scheme can improve timing uncertainties from $\pm80\Myr$ to $\pm3.3\Myr$.

@@

\begin{figure}[htb]
    \begin{center}
    \includegraphics[width=\columnwidth]{plots/bet-on-it/6_mergerfraction.png}
    \caption{The fraction of low-mass (green) and high-mass (pink) orbits that merge before z=0 shown as a function of redshift. 
    The horizontal line shows a merger fraction of 0.9, or 90\%. 
    High-mass pairs have merger fractions greater than 0.9 for $z>1.1$, and low-mass pairs for $z>1.6$.
    %At $z>1$, the fraction of selected pairs that merge before $z=0$ is upwards of 80\% for both low mass and high mass isolated pairs. 
    The sharp decline of the merger fraction to zero at low redshift is a non-physical feature of the simulation ending at $z=0$.
    Only orbits with sufficiently short merger timescales will merge at redshifts $z<\sim1$.}
    \label{fig:fmerge}
    \end{center}
\end{figure}

\subsection{Number of Pairs}\label{sec:pairprops-num}
The total number of pairs at a given redshift is equal to the number of orbits at that redshift, including both merger and non-merger pairs. 
%We calculate the total number of pairs by selecting all pairs at a given redshift, including merger and non-merger pairs.
A single, non-merging pair with first infall at $z=3$ will contribute to the number of pairs at all redshifts from $z=0-3$.

@@

As subhalos experience first infall into a FoF group, the number of pairs increases. 
On the other hand, mergers simultaneously lead to a decrease in the number of pairs.
% The difference between high and low mass
The number of low-mass pairs increases from $\sim600$ pairs at $z=6$ to $\sim6,600$ at $z=2$. 
The decrease in the number of pairs after $z=2$ means that the number of low-mass pairs that are merging at each redshift is larger than the rate at which pairs are added to the sample.

@@

\subsection{Merger Fraction}\label{sec:pairprops-frac}
    We calculate the merger fraction by dividing the number of pairs that merge (dashed lines in Fig.~\ref{fig:numorbits}) by the total number of merging and non-merging orbits (solid lines) at a given redshift. 
    As before, an orbit with first infall at $z=2$ and merger at $z=1$ will be included in the merger fraction calculation for redshifts $z=1-2$.
    Note, this definition of the merger fraction differs from that typically used in observational studies, where merger fractions are computed from close pair fractions with a correction term~\cite[e.g.,][]{Ventou2019}.

@@

\begin{figure*}[htb!]
    \begin{center}
    \includegraphics[width=0.7\textwidth]{plots/bet-on-it/8_2Dhist.png}
    \caption{The distribution of merger timescales for low-mass (top) and high-mass (bottom) pairs as a function of physical separation at $z=0.5,1,1.5,2,3,4,5\mbox{and }6$. 
    The colorbars show the percentage of the total number of pairs at each redshift (given in the bottom right corner of each panel) that are in a given separation bin.
    In each of the first panels, the grey horizontal dashed line shows the time remaining in the simulation, above which there are no mergers. 
    %
    We find:
    i.) a positive correlation between separation and the time until merger at each selected redshift -- the lowest separation pairs tend to have the shortest merger timescales; 
    ii.) given a redshift and separation, the low-mass merger timescale is longer than the high-mass merger timescale, i.e. low-mass pairs at $z=1$ with separations $\rsep<100\,\kpc$ merge within $6\, \Gyr$, while high-mass pairs in the same separation range merge within $3\, \Gyr$; 
    iii.) merger timescales are longer for lower redshift pairs; 
    iv.) the spread of merger timescales and separations increase with decreasing redshift. }
    \label{fig:2dhist}
    \end{center}
\end{figure*}

\section{Results: The Mass and Redshift Dependence of Merger Timescales}\label{sec:results}
    Using our sample of isolated low-mass and high-mass pairs in the TNG100 simulation, we calculate the merger timescales, or time until merger, for all of the merging pairs in sample. 
    The merger timescale of a pair is defined to be the amount of time that elapses between the redshift at which pairs are selected and the merger redshift.
    Only the merging pairs will be considered for the remainder of the analysis. 
    %\footnote{Since our orbits are only defined after infall, the merger timescales will likewise only be calculated for pairs that are post-infall at the given separation.}

@@

    We find that the merger timescale for low-mass and high-mass pairs is positively correlated with the separation of the pair at a given redshift. 
    The merger timescale also increases with decreasing redshift for a given separation.
    Pairs at $z=4$ with separations between $50-100\,\kpc$ tend to merge in less time than pairs with the same separations at $z=0.5$. 
    For the same physical separation, the merger timescale of high-mass pairs is shorter than for low-mass at each redshift. 
    % However, the slope of this correlation decreases at higher redshifts.

    In addition, the spreads in the merger timescales and pair separations are smaller at higher redshift, for both low- and high-mass pairs. 
    For example, the spread of merger timescales for low-mass pairs at $z=1$ (4273 pairs) goes from $0-8\,\Gyr$, while pairs at $z=4$ (2218 pairs) have merger timescales between $0-3\,\Gyr$. 
    Likewise, the spread of separations at $z=1$ is $0-200\kpc$, but at $z=4$ the spread is smaller, between $10-125\kpc$. 
    The high-mass pairs also have a larger spread in the distribution of merger timescales and separations at low redshift than at higher redshift.
    This is due to the growth of halos over time, and thus an increasing virial radius for FoF groups at lower redshifts.

    Fig.~\ref{fig:2dhist} can be used to estimate the merger timescale of an isolated pair at a given redshift. 
    For example, a low-mass pair at $z=2$ with $\rsep\sim 75\,\kpc$ will merge in $0-5\,\Gyr$, with a most likely time to merger of around $2\Gyr$.
    On the other hand, a high-mass pair at $z=2$ with $\rsep\sim 75\,\kpc$ will merge in $0-2\,\Gyr$, with a most likely merger timescale of around $0-1\Gyr$.

@@

    \subsubsection{Full sample}
        We calculate the merger timescales for pairs at each redshift, and quantify the median and spread on the merger timescale as a function of redshift in Fig.~\ref{fig:timescales}. 
        We include the full catalog of merging orbits at each redshift, only excluding separations $\rsep<10\kpc$ to limit the impact of subhalos becoming indistinguishable in the \subfind{} catalogs.% to calculate the merger timescale. 
        % why do we want to consider the entire population of halos? why interesting?

@@

        Figure~\ref{fig:timescales} also shows
        the 1st and 3rd quartiles are indicated with shaded regions. 
        % low vs high
        Low-mass merger timescales (green) and high-mass merger timescales (pink) are roughly equivalent at all redshifts, which implies that mergers do not proceed in fundamentally different ways at different mass scales. 
        We explore this point further in Sections~\ref{sec:results-phys} and~\ref{sec:results-scal}.

@@

        It is notable that such a simple scaling holds for a complex simulation like TNG100 where subhalos are far from isotropic spherical distributions of mass.
        It is particularly interesting as well that low-mass and high-mass pairs have the same merger timescales \textit{and} follow the $1/H(z)$ redshift scaling. 
        %We believe this is due to the 
        % Thus, the primary mechanism that determines merger timescales is gravity.... and not dynamical friction? is this true?

@@

    \subsubsection{Physical Separation Selected Pairs }
    \label{sec:results-phys}
        Since pair samples are typically picked via separation criteria in both simulations and observations, we study the impact of different separation-based selection criteria on inferred merger timescales.
        Separation selected samples have a lower separation criteria of $10\,\kpc$.
        This lower separation criteria is also commonly applied to observationally selected pairs in studies of merger fractions and merger rates~\citep[see e.g.][and observational studies cited therein]{Lotz2011,Snyder2017,Besla2018}.

        The first set of separation criteria selects only pairs at a given redshift that have physical 3D separations greater than $10\,\kpc$ and less than $[50, 70, 100, 150, 200, \mbox{and }300]\kpc$, yielding six pair subsamples. 
        Pairs that are selected via the $\rsep<50\kpc$ separation cut will be included in the $\rsep<70\kpc$ subsample, and so on.
        These separation criteria do not vary as a function of time or mass, and are applied equivalently to the low-mass and high-mass samples at all redshifts. 
        %The separation criteria are applied independently at each redshift.
        For example, an orbit in the separation bin $<50\,\kpc$ at $z=2$ will not necessarily be in that same bin at other redshifts.
